---
title: "Sampling crop counts in strip trials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling crop counts in strip trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5)
```

Plant establishment counts, weed counts and disease scores are measured by hand
at a few places in each strip, after the treatment has been applied. They are
responses, not covariates: each strip is the experimental unit, and the counts
estimate its mean. This article answers the questions a trial team asks before
going out with a quadrat:

1. How many counts per strip, and how large a quadrat?
2. Where in the strip?
3. How should the counts be analysed?

The answers come from the AAGI-CU-RD-OFE Milestone 5 study (simulated
250 m x 12 m strips, with the variation set from public establishment trials).

```{r}
library(ofeIntegrateR)
```

## 1. How many counts?

`count_precision()` simulates strips with a known mean and counts them. The
**95% margin** is how close the counted strip mean comes to the truth in 95%
of strips.

```{r}
count_precision(n_stations = c(5, 8, 12), n_quadrats = 2, quadrat_area = 0.25,
                crop = "wheat", n_sim = 200, seed = 1)
```

For wheat or barley, 8 stations x 2 quadrats of 0.25 m^2 (1 m of row at 25 cm)
gives the strip mean to within about +/-13--15%. Canola has fewer plants per
m^2 and needs about twice the counted area:

```{r}
count_precision(n_stations = 8, n_quadrats = 4, quadrat_area = 0.5,
                crop = "canola", n_sim = 200, seed = 1)
```

Weeds are patchy, so even many quadrats leave a wide margin:

```{r}
count_precision(n_stations = c(12, 20), n_quadrats = 4, quadrat_area = 1,
                crop = "weeds", n_sim = 200, seed = 1)
```

The patchiness of a strip is not known in advance. The presets cover the range
of the public data: `patchiness = "low"` (research plots) to `"high"`
(commercial seeder runs). In a patchy paddock, add stations:

```{r}
count_precision(n_stations = c(8, 12), n_quadrats = 2, quadrat_area = 0.25,
                crop = "wheat", patchiness = "high", n_sim = 200, seed = 1)
```

`count_sample_size()` turns a target into a design:

```{r}
count_sample_size(0.10, crop = "wheat", n_quadrats = 4, quadrat_area = 0.25,
                  max_stations = 20, n_sim = 200, seed = 1)
```

For the same number of quadrats, spreading them over more stations is as good
or better than more quadrats per station, especially for weeds.

## 2. Where?

`place_count_stations()` puts evenly spaced stations along every plot of a
trial layout, at the same positions in each plot, with buffers at the ends and
sides. The defaults (20 m and 2 m) avoid headland effects, the build-up of a
treatment at the start of a run, and spill-over from the neighbouring strip.

```{r}
d <- make_trial_design(c("Control", "High rate"), n_rep = 4,
                       plot_width = 12, plot_length = 250, cell_size = 12,
                       seed = 1)
st <- place_count_stations(d, n_stations = 8, n_quadrats = 2, seed = 1)
head(st)
plot(st)
```

The coordinates (`x`, `y`) can be loaded into a GPS or a field app. Counting
only near the end where the strip is entered doubled the error in the
simulations, and did the same on 7700 real 260 m transects of commercial fields
surveyed for black-grass, because weed patches there extend over 100--150 m.

## 3. Analyse strip means, not quadrats

`analyse_strip_counts()` pools the counts in each strip and compares
treatments on the log scale with replicate blocks. Here are counts simulated
for the plan above, with a 20% higher establishment under the high rate and
some strip-to-strip variation:

```{r}
set.seed(2)
strip_eff <- exp(rnorm(8, 0, 0.1))
true_dens <- 120 * ifelse(st$treat == "High rate", 1.2, 1) * strip_eff[st$plot]
st$count <- rpois(nrow(st), true_dens * 0.25)
fit <- analyse_strip_counts(st, count = "count", strip = "plot",
                            treat = "treat", block = "rep", area = 0.25)
fit
```

The test has 3 degrees of freedom (4 replicates), whatever the number of
quadrats. Treating the 128 quadrats as replicates instead would ignore that
quadrats in one strip share its seeder pass: in the Milestone 5 simulations
that produced "significant" differences in up to 70% of trials with no
treatment effect.

For weeds and disease, count the same pegged quadrats before and after the
treatment and pass the earlier counts as `baseline`; the response is then the
change in each strip, which removes most of the patchiness.

## What strip trials can and cannot show

The counts above describe each strip well. Whether a *treatment difference* in
establishment can be detected depends mostly on the number of replicate strips
and on how consistently the seeder sows from one pass to the next. In the
Milestone 5 simulations a 20% difference in wheat establishment was detected in
only about half of trials with 8 replicates; large weed and disease effects
(50% fewer weeds) were detected reliably with 4--6 replicates. If a 10--20%
difference in establishment matters, a small-plot trial is the better tool.
