---
title: "Integrating point-source data with a yield map"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating point-source data with a yield map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
set.seed(2026)
```

An on-farm experiment usually produces two kinds of data. A **dense layer** —
yield monitor, EM38, NDVI — covers the whole trial at low marginal cost but
measures the outcome rather than its cause. A **point-source layer** — soil
cores, tissue tests, disease counts — carries direct agronomic meaning but is
expensive, so a trial can afford only tens of observations.

This vignette walks through combining them to estimate a treatment effect.
Every step runs on open-source engines; ASReml-R is used automatically when
licensed but is never required.

```{r}
library(ofeIntegrateR)
```

## Before the season: where should the cores go?

The sample count fixes the cost, so the design question is where to put them.
`place_point_samples()` implements the common schemes.

```{r}
design_grid <- simulate_ofe_trial(n_row = 24, n_col = 12, n_point_samples = 5,
                                  seed = 1)$grid

spread <- function(p) round(mean(dist(cbind(p$col, p$row))), 1)
sapply(c("random", "grid", "stratified", "nested"),
       function(d) spread(place_point_samples(design_grid, n = 16, design = d)))
```

Higher numbers mean samples are further apart. Two points follow from the
simulation work behind this package:

- **Spread beats clustering when the samples will build a covariate surface.**
  A systematic `"grid"` or a `"stratified"` draw beats simple random placement,
  though the margin is second-order next to the sample count.
- **`"nested"` is a reconnaissance design, not an interpolation design.** It
  concentrates pairs at short lags, which is what identifies a variogram range.
  Reusing that geometry for the integration sample leaves much of the trial far
  from any observation and produces a worse covariate surface for the same money.

## The dense layer, as a real trial produces it

Yield-monitor data are not a tidy grid. They arrive as a cloud of
GPS-referenced pings along harvester passes, with position error, over a paddock
that is rarely a rectangle, and with the occasional pass missing.
`simulate_yield_monitor()` generates that shape, so the workflow below is the one
you would actually run.

```{r}
sim <- simulate_yield_monitor(n_point_samples = 30, seed = 11, defects = TRUE)

nrow(sim$cloud)
head(sim$cloud, 3)
sim$true_effects
```

The contrast we are trying to recover is **C − A = 1.6**.

`defects = TRUE` also gives the cloud the faults a real file has, and labels
them, so the next step can be checked rather than trusted:

```{r}
table(sim$cloud$defect, useNA = "no")
```

## Step 0 — clean the file

The opening metres of each pass read low while grain is still reaching the
sensor, the closing metres read high, the machine slows into every turn, and a
scatter of readings are zero or several times the crop.

```{r}
clean <- clean_yield_monitor(sim$cloud, pass = "pass", order = "seq",
                             speed = "speed", pass_trim = 12, local_mad = 3)
attr(clean, "cleaning")
```

It is tempting to skip this and let gridding average the noise away. That works
for the random outliers and not for the rest: the pass-end faults are
*systematic*, with the low readings at the start of every pass and the high ones
at the end, so averaging preserves them and hands the model a trend running
along the direction of travel.

Whether that trend reaches the treatment estimate depends on how the passes lie
against the strips. Here they run **along** the strips, so every pass sits
wholly inside one treatment and the fault falls equally on all of them: across
six simulated trials, cleaning moves the recovered contrast by about one
percentage point. Where passes run **across** the strips it does not fall
equally, and that is when the trend becomes a treatment effect. Clean either
way — the cost is small and you do not always control the direction of
harvest — but do not expect the contrast to move much when the geometry is
kind.

Because the simulator labelled what it damaged, the rules can be scored:

```{r}
truth <- ifelse(is.na(sim$cloud$defect), "clean", sim$cloud$defect)
audit <- clean_yield_monitor(sim$cloud, pass = "pass", order = "seq",
                             speed = "speed", pass_trim = 12, local_mad = 3,
                             drop = FALSE)
round(tapply(!audit$.keep, truth, mean), 2)
```

Most of the pass-end damage goes and very few sound points go with it. The
outliers fare worse: half of them are dropouts reading zero, which is not far
enough from the median to trip a dispersion threshold. On data in physical units
you would add `min_yield = 0`, which is off by default because the function
cannot see the units.

## Step 1 — put the dense layer on an estimable lattice

Spatial mixed models with a separable AR1 residual need a complete rectangular
lattice. `grid_dense_layer()` snaps the cloud onto one, averages within cells,
and — importantly — keeps cells that contain no observations as missing values
rather than dropping them.

```{r}
g <- grid_dense_layer(clean, response = "yield", treat = "treat",
                      cell_size = 9)

dim(g)
table(empty_cells = g$n_obs == 0)
summary(g$n_obs[g$n_obs > 0])
```

Two columns are worth checking before going further.

`n_obs` is how many pings support each cell. Cells resting on one or two noisy
pings can be blanked with `n_min`. `treat_purity` is the share of a cell's
observations belonging to its assigned treatment; cells straddling a strip
boundary are ambiguous and are usually excluded.

```{r}
table(pure = g$treat_purity == 1, useNA = "ifany")

# Exclude boundary cells, but keep the lattice complete
g$yield[!is.na(g$treat_purity) & g$treat_purity < 0.8] <- NA
```

## Step 2 — krige the point samples onto the same grid

A real trial samples wherever the sampler could reach, not at cell centres.
`krige_point_samples()` fits a variogram and predicts onto the grid, so the
sparse layer becomes a covariate defined everywhere.

```{r}
g$x <- g$x_centre
g$y <- g$y_centre
kr <- krige_point_samples(sim$point_samples, g, value = "point_obs",
                          coords = c("x", "y"))

attr(kr, "variogram")
```

Always look at the fitted variogram. A range far larger than the trial, or a
nugget close to the sill, means the surface is not being estimated.

### What scale does the sampled layer vary at?

Before kriging anything, look at its variogram. The range says how far one core
speaks for, and the nugget says how much variation no sampling density will
ever resolve.

```{r, fig.width = 7, fig.height = 4.2}
v <- ofe_variogram(sim$point_samples, value = "point_obs",
                   x = "x", y = "y")
v
plot(v)
```

Two things to check on the printed output. The **nugget ratio** classifies the
spatial dependence: a weak-dependence layer is noise at the distances sampled,
and kriging it will return something close to its mean whatever you do with it.
And if the fitted practical range is longer than the largest lag the samples
cover, the variogram never reached its sill inside the data and the range is an
extrapolation — it says so rather than reporting the number quietly.

The variogram feeds the sampling calculation directly, so the range is not
transcribed by hand — it is the number that matters most there, and the one a
variogram reports two ways:

```{r}
plan <- kriging_sample_interval(v, target_kse = 0.8, area_ha = 2.6)
c(floor = round(plan$kse_floor, 3), interval = round(plan$interval, 1),
  n = plan$n_samples)
```

`kse_floor` is usually the most useful number of the three. The nugget cannot
be filtered out at an unsampled point, so no sampling density whatever gets the
kriging standard error below it — here about `r round(plan$kse_floor, 2)` of the
field standard deviation. Asking for anything tighter is not a budget problem,
it is impossible, and the function says so rather than returning an interval.

### Is the point layer dense enough?

`cv_krige_surface()` leaves each sample out in turn and predicts it from the
rest.

```{r}
cv <- cv_krige_surface(sim$point_samples, value = "point_obs",
                       coords = c("x", "y"))
c(rmse = round(cv$rmse, 3), r2 = round(cv$r2, 3))
```

Read this as a **sampling** diagnostic. A negative `r2` means kriging predicts
worse than the sample mean — the layer is too sparse or too noisy to reconstruct
its own surface, and the answer is more samples next season. It does not by
itself condemn the current analysis, and a healthy `r2` does not guarantee the
surface is *relevant to yield*: a soil property can be mapped perfectly and still
explain nothing about the treatment response.

## Step 3 — compare the integrated fit against the free baseline

The estimate to beat is the dense layer analysed on its own with a spatial
residual. It costs nothing and requires no sampling, so it is the honest
comparator for anything that does. `compare_integration()` fits both and lines
them up.

```{r}
cmp <- compare_integration(kr, response = "yield", treat = "treat",
                           covariate = "point_obs_kriged")

cmp[grepl("treat", cmp$term), ]
```

Compare `baseline_estimate` and `integrated_estimate` against the truth of 1.6,
and look at `se_change` — negative means the integrated fit is more precise.
Report both models in your write-up: the difference between them is exactly what
the sampling bought.

The two fitted models are kept for inspection:

```{r}
names(attr(cmp, "models"))
```

To fit either model on its own, use `fit_integrated_kriged()` directly;
`covariate = NULL` gives the baseline.

## Integrating more than one point variable

Pass several kriged surfaces and they all enter the model:

```{r, eval = FALSE}
fit_integrated_kriged(kr, response = "yield", treat = "treat",
                      covariate = c("soil_n_kriged", "soil_p_kriged"))
```

## Choosing an engine

| `engine` | Spatial residual | Random effects | Licence |
|---|---|---|---|
| `"ofe"` (default) | `ar1(row):ar1(col)`, via `fit_ofe()` | yes | open source |
| `"asreml"` | `ar1(row):ar1(col)` | yes | commercial |
| `"lme"` | exponential, via `nlme::lme()` | yes | open source |
| `"gls"` | exponential, via `nlme::gls()` | no | open source |
| `"lm"` | none | no | open source |

`"ofe"` is the default: it fits the same separable AR1 model asreml does, in
base R. `"lm"` ignores spatial correlation entirely and is best kept as a
diagnostic reference rather than a final model. `extract_fixed_effects()`
returns the same tidy table whichever engine fitted the model, so downstream
code does not change when a licence appears or disappears.

## Step 4 — fit the model yourself

`fit_integrated_kriged()` is a convenience wrapper. When you want a residual
structure of your own, call `fit_ofe()` directly; it takes the asreml
spellings, so the formula below is the one you would write for
`asreml::asreml()`.

```{r}
fit <- fit_ofe(yield ~ treat + point_obs_kriged, data = kr,
               residual = ~ ar1(row):ar1(col))
summary(fit)
```

`wald_tests()` tests each fixed term as a whole — the question a trial report
asks — rather than one coefficient at a time:

```{r}
wald_tests(fit)
```

and `ofe_means()` gives the treatment means, or the contrasts between them with
the standard error of difference:

```{r}
ofe_means(fit, "treat")
ofe_means(fit, "treat", pairwise = TRUE)
```

`ofe_lsd()` puts those together into the table a report prints: means sorted
best-first, with the a/b/c letters. Levels sharing a letter are not separable
at `alpha`.

```{r}
tab <- ofe_lsd(fit, "treat")
tab
attr(tab, "lsd")
```

On a balanced design this reproduces `agricolae::LSD.test()` exactly, and
`adjust = "tukey"` reproduces `agricolae::HSD.test()`. The default here differs
from both in one respect that matters for a spatial model: each pair is judged
on its own standard error of difference, because neighbouring strips are
compared more precisely than distant ones. Pass `use = "lsd"` for the single
average LSD a published table usually means, and quote the value from
`attr(tab, "lsd")` with it.

The default `adjust = "none"` gives unprotected comparisons — the convention
these letters usually carry, and not a family-wise error rate. With more than
three or four treatments, say so in the caption or adjust:

```{r}
ofe_lsd(fit, "treat", adjust = "tukey")
```

### Did the spatial model absorb the pattern?

`plot()` on a fit gives four panels, and the first and last are the ones that
earn their place. A spatial model is fitted precisely to absorb the field's
pattern, so the question it has to answer is whether any is left: patches still
visible on the residual map, or a residual variogram still climbing at short
lags, mean the residual structure has not done its job and the treatment
standard errors are optimistic.

```{r, fig.width = 8, fig.height = 6.5}
plot(fit)
```

The residual map uses a diverging scale forced symmetric about zero, so
over- and under-prediction of the same size get the same weight of ink.

## Step 5 — pseudo-environments

A strip runs the length of the paddock, and the treatment effect at one end
need not be the effect at the other. `partition_pseudo_env()` finds the zones
from the data: it removes the treatment signal, collapses what is left to a
profile along the trial, and segments that profile by dynamic programming.

The guard against inventing zones is the spatial covariance. A smooth field
always looks like it has regions; what matters is whether a region is wider
than the correlation range, because anything narrower is one realisation of the
same correlated surface. The practical range of a fitted exponential variogram
therefore sets the minimum zone width, and caps how many zones are on offer.

```{r}
z <- partition_pseudo_env(kr, response = "yield", along = "row",
                          treat = "treat")
p <- attr(z, "partition")
c(range = p$range, min_width = p$min_zone_width,
  most_zones_possible = p$max_zones_possible, chosen = p$n_zones)
p$zones
```

On this trial it returns a single zone, which is the right answer: nothing
zone-like was simulated. The trial is only twelve rows deep and the fitted
range is about four, so at most three zones were ever on offer, and BIC
preferred none of them. A function that always found zones would be worse than
useless.

`adaptive_residual()` then writes the residual structure those zones imply.
It gives a zone `ar1()` only in the dimensions where it has extent: a zone one
row deep would otherwise ask for an unidentifiable parameter and fail.

```{r}
r <- adaptive_residual(z)
r
attr(r, "geometry")
```

If more than one zone was found, fitting a zone-specific treatment effect is
one call:

```{r}
if (nlevels(z$zone) > 1) {
  fit_z <- fit_ofe(yield ~ zone + zone:treat, data = z, residual = r)
  print(wald_tests(fit_z))
  # One lettering per zone: comparisons never cross a zone boundary, and
  # each zone gets its own LSD.
  print(ofe_lsd(fit_z, "treat", by = "zone"))
}
```

### Zoning from covariates instead, when you have them

`partition_pseudo_env()` works from the response and cuts across the trial, so
every treatment stays in every zone. When the zones should instead come from
what is known before harvest -- elevation, an EM38 or gamma survey, a soil test
grid -- and may be any shape, `partition_paddock()` clusters those covariates
directly. Here is a synthetic elevation and soil layer over the same grid:

```{r}
kr$elevation <- 100 + 4 * exp(-((kr$y_centre - 30)^2) / 400)
kr$clay <- ifelse(kr$x_centre > 120, 20, 32)
set.seed(3)
kr$elevation <- kr$elevation + rnorm(nrow(kr), 0, 0.2)
kr$clay <- kr$clay + rnorm(nrow(kr), 0, 1)

pz <- partition_paddock(kr, covariates = c("elevation", "clay"),
                        treat = "treat")
attr(pz, "partition")$zones
```

Every zone is a rectangle, and `x_min`...`y_max` are its corners, so a block
can be marked out in the paddock. That is what the default `method` buys:
cutting a rectangle all the way across leaves two rectangles, so no number of
cuts can produce a zone that sends a finger out between its neighbours.
`method = "skater"` relaxes that to contiguous-but-any-shape, which suits a
boundary running at an angle; `method = "kmeans"` drops contiguity altogether
and is worth running once to see why it is not the default. `patches` counts
the connected pieces of each zone -- 1 throughout here, more than 1 as soon as
k-means fragments.

The `treatments` column counts the treatment levels present in each zone.
Zones of arbitrary shape will often fail to contain every treatment, which is
exactly when `partition_pseudo_env()` is the right tool instead: a zone that
does not see all the treatments cannot support a `zone:treat` term.

Two cautions. Zones found this way are a description of where the paddock
differs, not proof of a boundary — a smooth field with no step in it is still
split more often than not. And because the cuts were placed where the residuals
already differed, a zone-by-treatment interaction estimated on self-derived
zones is exploratory: use the zones freely for the residual structure, and
confirm an interaction against an independent layer (elevation, EM38, last
season's yield) by passing that layer as `response` instead.

## When integration is not worth it

Simulation work behind this package (GRDC project AAGI-CU-RD-OFE, Milestone 4)
found the benefit is real but conditional. Three findings are worth carrying into
planning:

- **Thirty samples is a floor, not a target.** On idealised lattices thirty
  samples were enough. On realistic trial geometry the same thirty bought very
  little, because averaging many yield pings per cell already makes the baseline
  accurate — the tidy benchmark had understated the baseline, not overstated the
  method.
- **Short-range variation defeats it.** Where the point variable decorrelates
  over a distance comparable to the grid cell, no affordable number of samples
  reconstructs the surface. Check the fitted variogram range against your cell
  size before committing.
- **A significant covariate is not evidence that integration helped.** With
  hundreds of grid cells the kriged covariate clears `|t| > 2` almost
  automatically, including where it does nothing for the treatment estimate.
  Judge the method on how the treatment contrasts move, which is what
  `compare_integration()` reports.

## The joint bivariate model

`fit_integrated_joint()` models the dense and point layers together with a
shared spatial random effect, estimating their cross-covariance instead of
treating the kriged surface as known. It is the right choice when that
covariance is itself of interest, but it needs an unstructured cross-covariance
to inform the treatment effect at all, and that is hard to estimate from sparse
point data. For routine treatment-effect work the kriged-covariate route is the
dependable default.
