---
title: "Integrating point-source data with a yield map"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating point-source data with a yield map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
set.seed(2026)
```

An on-farm experiment usually produces two kinds of data. A **dense layer** —
yield monitor, EM38, NDVI — covers the whole trial at low marginal cost but
measures the outcome rather than its cause. A **point-source layer** — soil
cores, tissue tests, disease counts — carries direct agronomic meaning but is
expensive, so a trial can afford only tens of observations.

This vignette walks through combining them to estimate a treatment effect. Every
step runs on open-source engines; ASReml-R is used automatically when licensed
but is never required.

```{r}
library(ofeIntegrateR)
```

## The starting point: data as a real trial produces it

Yield-monitor data are not a tidy grid. They arrive as a cloud of
GPS-referenced pings along harvester passes, with position error, and the
paddock is rarely a rectangle. Here we simulate that shape so the workflow is
the one you would actually run.

```{r}
# Harvester passes 4 m apart, pings ~1 m apart along each pass
pass_y <- rep(seq(2, 58, by = 4), each = 100)
cloud <- data.frame(
  x = rep(seq(1, 99, length.out = 100), times = length(unique(pass_y))),
  y = pass_y + rnorm(length(pass_y), 0, 0.3)          # GPS jitter
)

# Three treatment strips across the paddock
cloud$treat <- cut(cloud$x, breaks = c(0, 33, 66, 100), labels = c("A", "B", "C"))

# A spatially structured soil property that partly drives yield
soil <- as.vector(scale(with(cloud, sin(x / 18) + cos(y / 12) + rnorm(nrow(cloud), 0, 0.3))))
cloud$yield <- c(A = 0, B = 0.8, C = 1.6)[as.character(cloud$treat)] +
  0.6 * soil + rnorm(nrow(cloud), 0, 0.4)

nrow(cloud)
head(cloud, 3)
```

The true treatment contrast we are trying to recover is **C − A = 1.6**.

## Step 1 — put the dense layer on an estimable lattice

Spatial mixed models with a separable AR1 residual need a complete rectangular
lattice. `grid_dense_layer()` snaps the cloud onto one, averages within cells,
and — importantly — keeps cells that contain no observations, as missing values
rather than dropped rows.

```{r}
g <- grid_dense_layer(cloud, response = "yield", treat = "treat", cell_size = 5)

dim(g)
table(empty_cells = g$n_obs == 0)
summary(g$n_obs[g$n_obs > 0])
```

Two columns are worth checking before going further.

`n_obs` is how many pings support each cell. Cells resting on one or two noisy
pings can be blanked with `n_min`:

```{r}
g_strict <- grid_dense_layer(cloud, response = "yield", treat = "treat",
                             cell_size = 5, n_min = 3)
sum(is.na(g_strict$yield))
```

`treat_purity` is the share of a cell's observations belonging to its assigned
treatment. Cells straddling a strip boundary are ambiguous and are usually
excluded:

```{r}
table(pure = g$treat_purity == 1, useNA = "ifany")

# Exclude boundary cells but keep the lattice complete
g$yield[!is.na(g$treat_purity) & g$treat_purity < 0.8] <- NA
```

## Step 2 — krige the point samples onto the same grid

A real trial samples wherever the sampler could reach, not at cell centres.
`krige_point_samples()` fits a variogram and predicts onto the grid, so the
sparse layer becomes a covariate defined everywhere.

```{r}
# 30 soil cores at arbitrary locations
idx <- sample(nrow(cloud), 30)
cores <- data.frame(x = cloud$x[idx], y = cloud$y[idx],
                    soil_n = soil[idx] + rnorm(30, 0, 0.3))

# Kriging works in the coordinate system of both layers
g$x <- g$x_centre
g$y <- g$y_centre
kr <- krige_point_samples(cores, g, value = "soil_n", coords = c("x", "y"))

names(kr)[grepl("soil_n", names(kr))]
attr(kr, "variogram")
```

Always look at the fitted variogram. A range far larger than the trial, or a
nugget close to the sill, means the surface is not being estimated and the
covariate will carry little information.

## Step 3 — fit the baseline first, then the integrated model

The estimate to beat is **the dense layer analysed on its own with a spatial
residual**. It costs nothing and requires no sampling, so it is the honest
comparator for anything that does.

```{r}
# covariate = NULL is the dense-layer-only baseline
baseline <- fit_integrated_kriged(kr, response = "yield", treat = "treat",
                                  covariate = NULL, engine = "gls")

integrated <- fit_integrated_kriged(kr, response = "yield", treat = "treat",
                                    covariate = "soil_n_kriged", engine = "gls")
```

`extract_fixed_effects()` gives a tidy table regardless of which engine fitted
the model, so downstream code does not need to change when a licence appears or
disappears.

```{r}
contrast_C <- function(fit) {
  fx <- extract_fixed_effects(fit)
  fx[grepl("treatC$", fx$term), c("term", "estimate", "se")]
}

rbind(baseline = contrast_C(baseline),
      integrated = contrast_C(integrated))
```

Compare both against the truth of 1.6. Report both in your write-up: the
difference between them is exactly what the sampling bought.

## Integrating more than one point variable

Pass several kriged surfaces and they all enter the model:

```{r, eval = FALSE}
fit_integrated_kriged(kr, response = "yield", treat = "treat",
                      covariate = c("soil_n_kriged", "soil_p_kriged"),
                      engine = "gls")
```

## Choosing an engine

| `engine` | Spatial residual | Licence |
|---|---|---|
| `"asreml"` | `ar1(row):ar1(col)` | commercial |
| `"gls"` | exponential, via `nlme::gls()` | open source |
| `"lm"` | none | open source |

`"gls"` is the open-source workhorse. `"lm"` ignores spatial correlation
entirely and is best kept as a diagnostic reference rather than a final model.

## When integration is not worth it

Simulation work behind this package (GRDC project AAGI-CU-RD-OFE, Milestone 4)
found the benefit is real but conditional. Three findings are worth carrying
into planning:

- **Sample count matters, and thirty is a floor rather than a target.** On
  idealised lattices thirty samples were enough; on realistic trial geometry the
  same thirty bought very little, because averaging many yield pings per cell
  already makes the baseline accurate.
- **Short-range variation defeats it.** Where the point variable decorrelates
  over a distance comparable to the grid cell, no affordable number of samples
  reconstructs the surface. Check the fitted variogram range against your cell
  size before committing.
- **A significant covariate is not evidence that integration helped.** With
  hundreds of grid cells the kriged covariate clears `|t| > 2` almost
  automatically. Judge the method by comparing treatment contrasts against the
  baseline, not by the covariate's p-value.

## The joint bivariate model

`fit_integrated_joint()` models the dense and point layers together with a
shared spatial random effect, estimating their cross-covariance instead of
treating the kriged surface as known. It is the right choice when that
covariance is itself of interest, but it needs an unstructured cross-covariance
to inform the treatment effect at all, and that is hard to estimate from sparse
point data. For routine treatment-effect work the kriged-covariate route is the
dependable default.
