---
title: "A worked example on real trial data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked example on real trial data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
has_agridat <- requireNamespace("agridat", quietly = TRUE)
knitr::opts_chunk$set(eval = has_agridat)
```

```{r, eval = !has_agridat, echo = FALSE, results = "asis"}
cat("This article needs the **agridat** package, which is not installed.",
    "Install it with `install.packages(\"agridat\")` to run the example.")
```

The other articles use simulated trials, where the truth is known. This one
uses a real trial, where it is not, and where the data arrive in the state real
data arrive in.

`agridat::lasrosas.corn` is an on-farm nitrogen strip trial from Las Rosas,
Córdoba, Argentina, recorded by a yield monitor in 1999 and 2001. It has what
this package is built for: GPS-referenced yield observations, a treatment
applied in strips across the paddock, and an ancillary spatial layer.

```{r}
library(ofeIntegrateR)
library(agridat)

d <- subset(lasrosas.corn, year == 1999)
str(d)
```

`yield` is in quintals per hectare, `nf` is the nitrogen rate (six levels),
`rep` the replicate, `bv` a brightness value from aerial imagery (a proxy for
soil organic matter), and `topo` one of four topographic zones.

## Project, then turn the field to face the strips

`grid_dense_layer()` works in the units of the coordinates it is given, so
degrees of latitude and longitude would give meaningless cell sizes. Project to
metres first. A local equirectangular approximation is accurate enough over a
single paddock; use a proper projection (`sf::st_transform()`) for anything
larger.

The strips here run at an angle to north. A lattice laid along north and east
would cut across them, so the coordinates are also rotated: `along` runs the
length of the strips and `across` runs over them.

```{r}
lat0 <- mean(d$lat)
lon0 <- mean(d$long)
d$x <- (d$long - lon0) * 111320 * cos(lat0 * pi / 180)
d$y <- (d$lat - lat0) * 110540

pc <- prcomp(cbind(d$x, d$y))
d$along  <- pc$x[, 1] - min(pc$x[, 1])
d$across <- pc$x[, 2] - min(pc$x[, 2])

round(c(length_m = diff(range(d$along)), width_m = diff(range(d$across))))
nrow(d)
```

About 1,700 yield readings over a field about 650 m long and 115 m wide.

## The layout: 18 strips, one harvester pass each

```{r}
strip <- interaction(d$nf, d$rep, drop = TRUE)
strip_mid <- sort(tapply(d$across, strip, median))
round(diff(strip_mid), 1)
names(strip_mid)
```

Six nitrogen rates in three reps make 18 strips. Each strip is a single
harvester pass, 6.7 m wide, running the full length of the field, and the same
order of rates is used in every rep. This is what the grid has to respect.

## Choosing a cell size is a real decision

The cell size trades two things against each other. Large cells average more
yield readings, which suppresses monitor noise. Small cells are less likely to
straddle a treatment boundary. `choose_cell_size()` reports both sides of that
trade-off.

```{r}
choose_cell_size(d, x = "along", y = "across", response = "yield", treat = "nf",
                 sizes = c(4, 5, 6, 8, 10))
```

`pure` is the share of occupied cells lying wholly inside one nitrogen rate. It
stays at 1 up to 6 m and collapses beyond that, because a cell wider than one
6.7 m strip must straddle two. `median_obs` stays at one reading per cell
throughout: this 1999 monitor logged a reading about every 6.7 m along a pass.
No size gives both several readings per cell and pure cells, and the warning
says so. That is a property of the data, not something a setting can fix.

The natural lattice for a strip trial puts one row of cells on each strip. With
cells 6.7 m square, centred on the passes, each lattice row is one strip.

```{r}
d$across <- d$across - strip_mid[1] + 6.7 / 2    # centre the first pass in a cell

g  <- grid_dense_layer(d, x = "along", y = "across", response = "yield",
                       treat = "nf", cell_size = 6.7)
gb <- grid_dense_layer(d, x = "along", y = "across", response = "bv",
                       cell_size = 6.7)
g$bv <- gb$bv[match(paste(g$row, g$col), paste(gb$row, gb$col))]

c(rows = max(g$row), cols = max(g$col), cells = nrow(g),
  with_yield = sum(!is.na(g$yield)))
```

Cells with no reading have no treatment label, but every row is a single strip,
so they take the rate of their row. Their yield stays missing; the label only
keeps the lattice complete for the spatial model.

```{r}
for (r in unique(g$row)) {
  in_row <- g$row == r
  g$nf[in_row & is.na(g$nf)] <- names(which.max(table(g$nf[in_row])))
}
g$yield[!is.na(g$treat_purity) & g$treat_purity < 0.8] <- NA
table(g$nf)
```

## Standing in for a coring campaign

This trial has no soil cores: `bv` was measured everywhere. To show the
integration workflow we treat it as if it had been *sampled* at twelve
locations, which is what a trial team can typically afford. Growers often take a
core at each end of a strip, so we take one core near each end of one strip per
nitrogen rate. The surface being kriged is real; the sparsity is imposed.

```{r}
can_core <- which(!is.na(g$bv))
idx <- c()
for (lev in levels(g$nf)) {
  rows_lev <- unique(g$row[g$nf == lev])
  r <- rows_lev[sample(length(rows_lev), 1)]
  cells <- can_core[g$row[can_core] == r]
  lo <- cells[g$x_centre[cells] <= quantile(g$x_centre[cells], 0.10)]
  hi <- cells[g$x_centre[cells] >= quantile(g$x_centre[cells], 0.90)]
  idx <- c(idx, lo[sample(length(lo), 1)], hi[sample(length(hi), 1)])
}

cores <- data.frame(x = g$x_centre[idx], y = g$y_centre[idx], bv_obs = g$bv[idx])
g$x <- g$x_centre
g$y <- g$y_centre
kr <- krige_point_samples(cores, g, value = "bv_obs", coords = c("x", "y"))
```

Are twelve cores enough to reconstruct the surface?

```{r}
cv <- cv_krige_surface(cores, value = "bv_obs", coords = c("x", "y"))
round(c(rmse = cv$rmse, r2 = cv$r2), 2)
```

A positive `r2` would mean kriging beats simply using the average of the
cores. Here it is `r round(cv$r2, 2)`: `r if (cv$r2 > 0) "the cores do reconstruct some of the surface" else "twelve cores bunched at the two ends of the field cannot reconstruct the surface in between, and the kriged map is worse than a flat average"`.
Even a positive value would only say the surface is *estimated*, not that it is
*relevant to yield*.

## Baseline against integrated

```{r}
cmp <- compare_integration(kr, response = "yield", treat = "nf",
                           covariate = "bv_obs_kriged")
cmp[grepl("^nf|bv", cmp$term), ]
```

The nitrogen response is clear in both models: the contrasts rise from N1 to N5,
recovered from irregular yield-monitor data that started as a GPS point cloud.

It is tempting to read `se_change` as the value of the cores. Resist it. The
two models estimate their own spatial correlation, and a smooth kriged
covariate changes what the residual structure has left to explain, so the
reported standard errors can shrink or grow for reasons that have nothing to do
with accuracy. Here they shrank even though the twelve cores could not
reconstruct the soil surface at all. In the analysis behind the table below,
repeating the twelve-core draw gave standard errors that moved up and down from
one draw to the next, sometimes below those from the full `bv` survey, which no
twelve cores can genuinely beat. Comparing standard errors from two different
models is not a reliable score.

## Scoring against a known answer

A fairer test keeps the real field but puts back a known answer:

1. Fit the trial once with `bv` from the full survey and subtract the estimated
   nitrogen effects from every yield value. What remains is the real yield
   pattern and the real link between `bv` and yield.
2. Randomly reassign the six rates to the six strips within each rep, and add
   known effects back.
3. Analyse the result with and without the cores, and score each estimate
   against the known effects.

One round looks like this (not run here, because repeating it hundreds of times
takes a while):

```{r, eval = FALSE}
# step 1: remove the estimated N effects (bv from the full survey)
full <- krige_point_samples(d[, c("along", "across", "bv")],
                            transform(g, along = x_centre, across = y_centre),
                            value = "bv", coords = c("along", "across"))
fit_ref <- fit_integrated_kriged(full, response = "yield", treat = "nf",
                                 covariate = "bv_kriged")
fx  <- extract_fixed_effects(fit_ref)
tau <- c(0, fx$estimate[grepl("^nf", fx$term)])
names(tau) <- levels(g$nf)
g$yield0 <- g$yield - tau[as.character(g$nf)]

# step 2: a new random layout within each rep, known effects added back
rep_of_row <- tapply(as.character(d$rep),
                     floor((d$across - min(d$across)) / 6.7) + 1,   # = g$row
                     function(z) z[1])
new_nf <- character(max(g$row))
for (r in unique(rep_of_row)) {
  rows_r <- as.integer(names(rep_of_row)[rep_of_row == r])
  new_nf[rows_r] <- sample(levels(g$nf), length(rows_r))
}
g$nf_k    <- factor(new_nf[g$row], levels = levels(g$nf))
g$yield_k <- g$yield0 + tau[as.character(g$nf_k)]

# step 3: analyse with and without the cores, score against tau
kr$nf_k <- g$nf_k
kr$yield_k <- g$yield_k
base <- fit_integrated_kriged(g,  response = "yield_k", treat = "nf_k")
with <- fit_integrated_kriged(kr, response = "yield_k", treat = "nf_k",
                              covariate = "bv_obs_kriged")
err <- function(m) {
  fx <- extract_fixed_effects(m)
  mean(abs(fx$estimate[grepl("^nf_k", fx$term)] - tau[-1]))
}
c(baseline = err(base), with_cores = err(with))
```

The package ships the result of 200 rounds per year, computed this way (with
rep as a fixed effect, a lattice of 10 m cells along each strip, and
ASReml-R), together with 12 random cores, 30 random cores, the full `bv` survey
and the four topographic zones for comparison:

```{r}
known <- read.csv(system.file("extdata", "lasrosas_known_answer.csv",
                              package = "ofeIntegrateR"))
known
```

`median_error` is the typical error in the five N-vs-N0 effects (q/ha),
`change` is relative to no soil information (negative is better), and
`better_than_none` is the share of rounds in which the analysis beat the one
with no soil information.

```{r, include = FALSE}
k99 <- known[known$year == 1999, ]
k01 <- known[known$year == 2001, ]
cores_rows <- grepl("cores", known$analysis)
pct <- function(x) paste0(round(100 * x), "%")
```

Three things stand out:

- **A dozen cores did little.** In 1999 they changed the error by
  `r pct(min(k99$change[grepl("cores", k99$analysis)]))` to
  `r pct(max(k99$change[grepl("cores", k99$analysis)]))`; in 2001 they made it
  slightly worse. They beat no soil information in only
  `r pct(min(known$better_than_none[cores_rows]))`--`r pct(max(known$better_than_none[cores_rows]))`
  of rounds.
- **Even the full survey added modestly.** Complete `bv` coverage cut the error
  by `r pct(-k99$change[k99$analysis == "Full bv survey"])` in 1999 and
  `r pct(-k01$change[k01$analysis == "Full bv survey"])` in 2001. Each strip runs
  the full length of the field and crosses every kind of soil, and the spatial
  model absorbs most of what is left, so the treatment comparison is already well
  protected.
- **Free zones were no substitute.** The topographic zones helped in 1999 and
  hurt in 2001.

This is one field and two seasons, and `bv` is only a proxy for soil, so treat
it as an illustration rather than a rule. It does show why the simulation
results in the other articles should be read as a best case.

## What to take from this

- Project your coordinates, and align the lattice with the strips before
  gridding.
- Choose the cell size from `n_obs` and `treat_purity`, not from convention.
  For a strip trial, one row of cells per strip is the natural choice.
- A handful of cores rarely sharpens a strip-trial estimate much, even when the
  soil layer is strongly related to yield.
- Do not judge a covariate by comparing standard errors from two models; score
  it against a known answer where you can.
