---
title: "A worked example on real trial data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked example on real trial data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
has_agridat <- requireNamespace("agridat", quietly = TRUE)
knitr::opts_chunk$set(eval = has_agridat)
```

```{r, eval = !has_agridat, echo = FALSE, results = "asis"}
cat("This article needs the **agridat** package, which is not installed.",
    "Install it with `install.packages(\"agridat\")` to run the example.")
```

The other article uses simulated trials, where the truth is known. This one uses
a real one, where it is not — and where the data arrive in the state real data
arrive in.

`agridat::lasrosas.corn` is an on-farm nitrogen experiment from Las Rosas,
Córdoba, Argentina, recorded by a yield monitor. It has what this package is
built for: GPS-referenced yield observations, a categorical treatment applied
across the paddock, and an ancillary spatial layer.

```{r}
library(ofeIntegrateR)
library(agridat)

d <- subset(lasrosas.corn, year == 1999)
str(d)
```

`yield` is in quintals per hectare, `nf` is the nitrogen treatment (six levels),
`bv` is a brightness value from aerial imagery, and `topo` is a topographic
class.

## Project the coordinates first

`grid_dense_layer()` works in the units of the coordinates it is given, so
degrees of latitude and longitude will produce meaningless cell sizes. Project
to metres before gridding. A local equirectangular approximation is accurate
enough over a single paddock; use a proper projection (`sf::st_transform()`) for
anything larger.

```{r}
lat0 <- mean(d$lat)
lon0 <- mean(d$long)
d$x <- (d$long - lon0) * 111320 * cos(lat0 * pi / 180)
d$y <- (d$lat - lat0) * 110540

round(c(width_m = diff(range(d$x)), height_m = diff(range(d$y))))
nrow(d)
```

About 1,700 yield observations over a 660 m × 240 m block.

## Choosing a cell size is a real decision

The cell size trades two things against each other. Large cells average more
yield observations, which suppresses monitor noise. Small cells are less likely
to straddle a treatment boundary. `grid_dense_layer()` reports both sides of
that trade-off, so it can be made on evidence rather than habit.

```{r}
sizes <- c(5, 8, 10, 15, 20)
do.call(rbind, lapply(sizes, function(cs) {
  g <- grid_dense_layer(d, x = "x", y = "y", response = "yield", treat = "nf",
                        cell_size = cs)
  data.frame(cell_m = cs,
             cells = nrow(g),
             with_data = sum(g$n_obs > 0),
             median_obs = median(g$n_obs[g$n_obs > 0]),
             pure = round(mean(g$treat_purity == 1, na.rm = TRUE), 2))
}))
```

The collapse in purity above 8 m is informative: it says the treatment units in
this trial are narrow. They are — all six nitrogen levels appear across the full
extent of the block, so this is a randomised trial in small plots rather than a
strip trial with wide contiguous strips.

That points to a cell size of 5–8 m. We take 8 m and exclude the cells that
still straddle a boundary.

```{r}
g <- grid_dense_layer(d, x = "x", y = "y", response = "yield", treat = "nf",
                      cell_size = 8)

# Carry the brightness layer onto the same lattice
gb <- grid_dense_layer(d, x = "x", y = "y", response = "bv", cell_size = 8)
g$bv <- gb$bv[match(paste(g$row, g$col), paste(gb$row, gb$col))]

# Exclude ambiguous cells, keeping the lattice complete
g$yield[!is.na(g$treat_purity) & g$treat_purity < 0.8] <- NA

c(cells = nrow(g), usable = sum(!is.na(g$yield)))
```

## Standing in for a sparse sampling campaign

This trial has no soil-core layer — `bv` was measured everywhere. To show the
integration workflow we treat it as if it had been *sampled* at forty locations,
the kind of budget a soil-coring campaign would have. This is a demonstration
device: the surface being kriged is real, but the sparsity is imposed.

```{r}
obs <- which(!is.na(g$bv) & !is.na(g$yield))
s <- sample(obs, 40)

cores <- data.frame(x = g$x_centre[s], y = g$y_centre[s], bv_obs = g$bv[s])

g$x <- g$x_centre
g$y <- g$y_centre
kr <- krige_point_samples(cores, g, value = "bv_obs", coords = c("x", "y"))
```

Is forty enough to reconstruct the surface?

```{r}
cv <- cv_krige_surface(cores, value = "bv_obs", coords = c("x", "y"))
round(c(rmse = cv$rmse, r2 = cv$r2), 2)
```

A positive `r2` means kriging beats the sample mean, so the layer is worth
carrying forward. Recall that this says the surface is *well estimated*, not
that it is *relevant to yield*.

## Baseline against integrated

```{r}
cmp <- compare_integration(kr, response = "yield", treat = "nf",
                           covariate = "bv_obs_kriged")

cmp[grepl("^nf|bv", cmp$term), ]
```

Two things to read here.

**The nitrogen response is clean.** The contrasts rise monotonically from N1 to
N5 — a textbook response curve, recovered from irregular yield-monitor data that
started as a GPS point cloud.

**Integration bought a real gain.** `se_change` is negative throughout, worth
between 6 and 11 per cent of the standard error. On this trial the forty-point
brightness surface did improve the treatment estimates.

## The baseline is only as strong as the model you give it

That second conclusion depends on which residual structure the comparison was
run under, and it is worth seeing why. Fit the same baseline — no covariate —
with the exponential-correlation engine instead, and compare the standard
errors it reports:

```{r}
base_gls <- fit_integrated_kriged(kr, response = "yield", treat = "nf",
                                  covariate = NULL, engine = "gls")
se <- data.frame(
  ar1 = cmp$baseline_se[grepl("^nf", cmp$term)],
  gls = extract_fixed_effects(base_gls)$se[-1]
)
round(cbind(se, ratio = se$gls / se$ar1), 3)
```

The exponential fit reports standard errors around half the size. On this
irregular, gappy lattice it does not absorb as much of the spatial structure,
so what it fails to model stays in the fit as apparent precision. Run
`compare_integration()` under that baseline and integration looks worthless —
not because the covariate stopped helping, but because an over-confident
baseline leaves nothing visible to improve on.

The lesson is not that one engine is right and the other wrong, but that
"integration did not help" is a claim about a *pair* of models. Before
concluding a covariate was not worth collecting, check that the baseline it was
judged against was itself well specified — `plot()` on the baseline fit, and its
residual variogram in particular, is the quickest way to see whether it was.

## What to take from this

- Project your coordinates before gridding.
- Choose the cell size from `n_obs` and `treat_purity`, not from convention.
  Their trade-off is specific to your trial's plot geometry.
- Expect integration to help when the baseline is weak, and to do little when it
  is already strong. Report both so the difference is visible.
- "Integration did not help" is a statement about two models, not one. Check the
  baseline is well specified before believing it.
