---
title: "A worked example on real trial data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked example on real trial data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
has_agridat <- requireNamespace("agridat", quietly = TRUE)
knitr::opts_chunk$set(eval = has_agridat)
```

```{r, eval = !has_agridat, echo = FALSE, results = "asis"}
cat("This article needs the **agridat** package, which is not installed.",
    "Install it with `install.packages(\"agridat\")` to run the example.")
```

The other article uses simulated trials, where the truth is known. This one uses
a real one, where it is not — and where the data arrive in the state real data
arrive in.

`agridat::lasrosas.corn` is an on-farm nitrogen experiment from Las Rosas,
Córdoba, Argentina, recorded by a yield monitor. It has what this package is
built for: GPS-referenced yield observations, a categorical treatment applied
across the paddock, and an ancillary spatial layer.

```{r}
library(ofeIntegrateR)
library(agridat)

d <- subset(lasrosas.corn, year == 1999)
str(d)
```

`yield` is in quintals per hectare, `nf` is the nitrogen treatment (six levels),
`bv` is a brightness value from aerial imagery, and `topo` is a topographic
class.

## Project the coordinates first

`grid_dense_layer()` works in the units of the coordinates it is given, so
degrees of latitude and longitude will produce meaningless cell sizes. Project
to metres before gridding. A local equirectangular approximation is accurate
enough over a single paddock; use a proper projection (`sf::st_transform()`) for
anything larger.

```{r}
lat0 <- mean(d$lat)
lon0 <- mean(d$long)
d$x <- (d$long - lon0) * 111320 * cos(lat0 * pi / 180)
d$y <- (d$lat - lat0) * 110540

round(c(width_m = diff(range(d$x)), height_m = diff(range(d$y))))
nrow(d)
```

About 1,700 yield observations over a 660 m × 240 m block.

## Choosing a cell size is a real decision

The cell size trades two things against each other. Large cells average more
yield observations, which suppresses monitor noise. Small cells are less likely
to straddle a treatment boundary. `grid_dense_layer()` reports both sides of
that trade-off, so it can be made on evidence rather than habit.

```{r}
sizes <- c(5, 8, 10, 15, 20)
do.call(rbind, lapply(sizes, function(cs) {
  g <- grid_dense_layer(d, x = "x", y = "y", response = "yield", treat = "nf",
                        cell_size = cs)
  data.frame(cell_m = cs,
             cells = nrow(g),
             with_data = sum(g$n_obs > 0),
             median_obs = median(g$n_obs[g$n_obs > 0]),
             pure = round(mean(g$treat_purity == 1, na.rm = TRUE), 2))
}))
```

The collapse in purity above 8 m is informative: it says the treatment units in
this trial are narrow. They are — all six nitrogen levels appear across the full
extent of the block, so this is a randomised trial in small plots rather than a
strip trial with wide contiguous strips.

That points to a cell size of 5–8 m. We take 8 m and exclude the cells that
still straddle a boundary.

```{r}
g <- grid_dense_layer(d, x = "x", y = "y", response = "yield", treat = "nf",
                      cell_size = 8)

# Carry the brightness layer onto the same lattice
gb <- grid_dense_layer(d, x = "x", y = "y", response = "bv", cell_size = 8)
g$bv <- gb$bv[match(paste(g$row, g$col), paste(gb$row, gb$col))]

# Exclude ambiguous cells, keeping the lattice complete
g$yield[!is.na(g$treat_purity) & g$treat_purity < 0.8] <- NA

c(cells = nrow(g), usable = sum(!is.na(g$yield)))
```

## Standing in for a sparse sampling campaign

This trial has no soil-core layer — `bv` was measured everywhere. To show the
integration workflow we treat it as if it had been *sampled* at forty locations,
the kind of budget a soil-coring campaign would have. This is a demonstration
device: the surface being kriged is real, but the sparsity is imposed.

```{r}
obs <- which(!is.na(g$bv) & !is.na(g$yield))
s <- sample(obs, 40)

cores <- data.frame(x = g$x_centre[s], y = g$y_centre[s], bv_obs = g$bv[s])

g$x <- g$x_centre
g$y <- g$y_centre
kr <- krige_point_samples(cores, g, value = "bv_obs", coords = c("x", "y"))
```

Is forty enough to reconstruct the surface?

```{r}
cv <- cv_krige_surface(cores, value = "bv_obs", coords = c("x", "y"))
round(c(rmse = cv$rmse, r2 = cv$r2), 2)
```

A positive `r2` means kriging beats the sample mean, so the layer is worth
carrying forward. Recall that this says the surface is *well estimated*, not
that it is *relevant to yield*.

## Baseline against integrated

```{r}
cmp <- compare_integration(kr, response = "yield", treat = "nf",
                           covariate = "bv_obs_kriged", engine = "gls")

cmp[grepl("^nf|bv", cmp$term), ]
```

Two things to read here.

**The nitrogen response is clean.** The contrasts rise monotonically from N1 to
N5 — a textbook response curve, recovered from irregular yield-monitor data that
started as a GPS point cloud.

**Integration changed almost nothing.** The treatment estimates barely move and
`se_change` is negligible. On this trial, adding a forty-point brightness surface
did not improve the treatment estimates, even though the surface itself was
adequately estimated.

That is not a failure of the method; it is the method working as intended. The
dense yield layer, averaged over several observations per cell and spread across
hundreds of cells, already estimates these contrasts well. A covariate has to
earn its place against that, and here it did not. Simulation work behind this
package found the same pattern whenever the baseline is strong — which is why
`compare_integration()` reports both models rather than the integrated one
alone.

## What to take from this

- Project your coordinates before gridding.
- Choose the cell size from `n_obs` and `treat_purity`, not from convention.
  Their trade-off is specific to your trial's plot geometry.
- Expect integration to help when the baseline is weak, and to do little when it
  is already strong. Report both so the difference is visible.
