---
title: "Worked examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
has_asreml <- requireNamespace("asreml", quietly = TRUE)
```

Four short recipes, each answering one question a trial team asks. They use
simulated data so that the true answer is known and the output can be checked
against it. `vignette("ofe-integration")` walks through the whole pipeline in
more depth.

```{r}
library(ofeIntegrateR)
```

## 1. A strip trial with a yield map and nothing else

*Question: what are the treatment means, and which differ?*

A yield-monitor file needs cleaning and gridding before it can be modelled.
Then the spatial model is fitted to the yield map alone. This is also the free
baseline that any point-sampling campaign has to beat.

```{r recipe1}
ym <- simulate_yield_monitor(treat_effects = c(0, 0.4, 0.8), defects = TRUE,
                             n_point_samples = 0, seed = 1)

# 1. Remove the points the harvester recorded but did not measure
clean <- clean_yield_monitor(ym$cloud, yield = "yield", pass = "pass",
                             speed = "speed")
attr(clean, "cleaning")

# 2. Snap the cloud onto a lattice; drop cells that straddle two treatments
g <- grid_dense_layer(clean, response = "yield", treat = "treat", cell_size = 9)
g$yield[!is.na(g$treat_purity) & g$treat_purity < 0.8] <- NA

# 3. Spatial model: treatment + AR1 x AR1 residual (no asreml licence needed)
fit <- fit_ofe(yield ~ treat, data = g, residual = ~ ar1(row):ar1(col))
wald_tests(fit)
ofe_lsd(fit, "treat")
```

The true effects were 0, 0.4 and 0.8 relative to treatment A:

```{r recipe1-check}
ofe_means(fit, "treat", pairwise = TRUE)[, c("contrast", "estimate", "std.error")]
```

## 2. Planning a soil-sampling campaign

*Question: how far apart should the cores be, and how many do we need?*

A variogram from a reconnaissance survey, or from ancillary data, sets the
sampling interval. `kriging_sample_interval()` finds the largest grid spacing
at which kriging reaches a target precision.

```{r recipe2}
trial <- simulate_ofe_trial(n_row = 40, n_col = 24, point_range = 8,
                            n_point_samples = 150, seed = 2)
# Work in metres: here each grid cell is 10 m, so the trial is 240 m x 400 m
recon <- transform(trial$point_samples, x = col * 10, y = row * 10)
v <- ofe_variogram(recon, value = "point_obs", x = "x", y = "y")
v$practical_range     # metres
v$range_identified    # is the range inside the distances actually sampled?

plan <- kriging_sample_interval(v, target_kse = 0.5, area_ha = 24 * 40 * 100 / 1e4)
plan$interval         # metres between cores
plan$n_samples
plan$kse_floor
```

The interval and the area must be in the same units, so give the
coordinates in metres before fitting the variogram.

Treat a variogram from one field with caution. When the range is not small
compared with the field, the range estimate varies a lot from one field to
the next, and `range_identified` reports whether the fitted range falls
inside the distances actually sampled.

`kse_floor` is the best relative precision that any sampling effort can
reach: at an unsampled location the nugget cannot be interpolated away. If the
target is below the floor, no interval meets it.

A note on units. In gstat, the `range` of an exponential model is a
*parameter*, and the correlation falls to 5% at about three times that
distance. `simulate_ofe_trial(point_range = 8)` therefore gives a practical
range of about 24 cells. `ofe_variogram()` reports both (`range` and
`practical_range`), and `kriging_sample_interval()` takes the practical range.

Place the cores to cover the trial rather than in clusters:

```{r recipe2-place}
set.seed(2)
cores <- place_point_samples(trial$grid, n = plan$n_samples, design = "grid")
nrow(cores)
```

## 3. A sparse response with a dense covariate: hand cuts and NDVI

*Question: we took four hand cuts per strip and flew NDVI over the whole
trial. Can NDVI make the hand-cut comparison more precise?*

This is where the joint bivariate model helps. The hand cuts and NDVI are
fitted as two correlated traits, each with its own treatment effects, so the
cuts borrow strength from NDVI through their residual correlation.

```{r recipe3-data}
p <- simulate_paddock(n_row = 24, n_col = 18, treatments = c("N0", "N60", "N120"),
                      n_rep = 3, treat_effect = c(0, 0.3, 0.5), seed = 3)
set.seed(3)
# NDVI follows the yield potential of the paddock and also responds to N
p$ndvi <- 0.55 + 0.06 * as.numeric(scale(p$yield_potential)) +
  c(N0 = 0, N60 = 0.03, N120 = 0.05)[as.character(p$treat)] +
  rnorm(nrow(p), 0, 0.02)
# Four hand cuts per strip
cuts <- do.call(rbind, lapply(split(p, p$plot), function(d) d[sample(nrow(d), 4), ]))
cuts <- data.frame(row = cuts$row, col = cuts$col, rep = cuts$rep,
                   treat = cuts$treat, cut_yield = cuts$yield,
                   ndvi = cuts$ndvi)
nrow(cuts)
```

The cuts alone (true N120 $-$ N0 effect: 0.5):

```{r recipe3-alone}
alone <- lm(cut_yield ~ factor(rep) + treat, data = cuts)
summary(alone)$coefficients[c("treatN60", "treatN120"), 1:2]
```

With NDVI, through the joint model:

```{r recipe3-joint, eval = has_asreml, warning = FALSE}
joint <- fit_integrated_joint(p[, c("row", "col", "treat", "rep", "ndvi")],
                              cuts[, c("row", "col", "cut_yield")],
                              response_dense = "ndvi",
                              response_point = "cut_yield", block = "rep")
joint$converge
fx <- extract_fixed_effects(joint)
fx[grepl("cut_yield\\):treat_N", fx$term), ]
```

The joint model needs an ASReml-R licence when one layer is sparse. The
open-source `sommer` engine fills missing responses with their median, which
would swamp the cuts, so it is refused for a sparse layer.

**What not to do.** NDVI responds to nitrogen, so entering it as an ordinary
covariate removes part of the treatment effect being estimated:

```{r recipe3-pitfall}
wrong <- lm(cut_yield ~ factor(rep) + treat + ndvi, data = cuts)
summary(wrong)$coefficients[c("treatN60", "treatN120"), 1:2]
```

The joint model avoids this because NDVI has its own treatment effects, so the
cuts are conditioned only on NDVI's deviations from its treatment means.

## 4. Check the pipeline against a known truth before trusting it

*Question: does this analysis recover the effect on a paddock like ours?*

Simulate a few paddocks with the structure you expect, run the same analysis,
and see whether the intervals cover the true effect. Here the truth is an
exponential field plus noise, while the model assumes AR1 $\times$ AR1, as it
would in practice.

```{r recipe4}
check_one <- function(seed) {
  p <- simulate_paddock(n_row = 20, n_col = 18, treatments = c("A", "B", "C"),
                        n_rep = 3, treat_effect = c(0, 0.2, 0.4), seed = seed)
  p$rep <- factor(p$rep)
  f <- fit_ofe(yield ~ treat, random = ~ rep, data = p,
               residual = ~ ar1(row):ar1(col))
  ca <- subset(ofe_means(f, "treat", pairwise = TRUE), contrast == "C - A")
  data.frame(seed = seed, estimate = ca$estimate, se = ca$std.error,
             covers_truth = abs(ca$estimate - 0.4) <= 1.96 * ca$std.error)
}
do.call(rbind, lapply(1:5, check_one))
```

Five paddocks are enough to catch a gross problem. To estimate coverage or
power, run a few hundred; `validation/validate.R` in the package source does
this for the package's own functions.
